---
title: "Lab 8: Efficiency and Parallelization"
author: "Faith Okaalo"
date: "03/15/2026"
format: html
editor: visual
---

### Problem 1: Vectorization

1.  This function generates an `n x k` dataset with all its entries drawn from a Poission distribution with mean `lambda`.

```{r}
# Original slow function
fun1 <- function(n = 100, k = 4, lambda = 4) {
  x <- NULL
  for (i in 1:n){
    x <- rbind(x, rpois(k, lambda))
  }
  return(x)
}

# Faster version (creates variables at once and shapes into matrix)
fun1alt <- function(n = 100, k = 4, lambda = 4) {
  matrix(rpois(n * k, lambda), nrow = n, ncol = k)
}
```

Show that `fun1alt` generates a matrix with the same dimensions as `fun1` and that the values inside the two matrices follow similar distributions. Then check, using `microbenchmark::microbenchmark`, the speed of the two functions:

```{r}
# Check same dimensions
set.seed(42)
r1 <- fun1()
r2 <- fun1alt()
dim(r1) == dim(r2)       
```

```{r}
# Check distributions
mean(r1); mean(r2)       

# Benchmark
library(microbenchmark)
microbenchmark(fun1(), fun1alt(), times = 100)
```

2.  This function finds the maximum value of each column of a matrix. Show that both functions return the same output for a given input matrix, `x`. Then check the speed of the two functions, as you did above

    ```{r}
    # Data generating process (10 x 10,000 matrix) *Original slow function
    set.seed(1234)
    x <- matrix(rnorm(1e4), nrow = 10)

    # Find each column's max value
    fun2 <- function(x) {
      apply(x, 2, max)
    }


    # # Faster version *matrix indexing instead of apply
    fun2alt <- function(x) {
      x[cbind(max.col(t(x)), seq_len(ncol(x)))]
    }

    # Check the output
    all(fun2(x) == fun2alt(x))   

    # Benchmark
    microbenchmark(fun2(x), fun2alt(x), times = 100)

    ```

### Problem 2: Parallelization

We will now turn our attention to the statistical concept of [bootstrapping](https://en.wikipedia.org/wiki/Bootstrapping_(statistics)). Among its many uses, non-parametric bootstrapping allows us to obtain confidence intervals for parameter estimates without relying on parametric assumptions. Don’t worry if these concepts are unfamiliar, we only care about the computation methods in this lab, not the statistics. The main assumption is that we can approximate the results of many repeated experiments by resampling observations from our original dataset, which reflects the population.

1.  This function implements a serial version of the bootstrap. Edit this function to parallelize the `lapply` loop, using whichever method you prefer. Rather than specifying the number of cores to use, use the number given by the `ncpus` argument, so that we can test it with different numbers of cores later.

    ```{r}
    my_boot <- function(dat, stat, R, ncpus = 1L) {
      
      # Get random row indices for each bootstrap sample
      n <- nrow(dat)
      idx <- matrix(sample.int(n, n * R, TRUE), nrow = n, ncol = R)
      
      # If more than 1 core, run in parallel, otherwise run normally
      if (ncpus > 1L) {
        cl <- makeCluster(ncpus)                                     
        clusterExport(cl, varlist = c("dat", "idx", "stat"),    
                      envir = environment())
        ans <- parLapply(cl, seq_len(R), function(i) {      
          stat(dat[idx[, i], , drop = FALSE])
        })
        stopCluster(cl)                                    
      } else {
        ans <- lapply(seq_len(R), function(i) {                         
          # regular lapply (1 core)
          stat(dat[idx[, i], , drop = FALSE])
        })
      }
       
    # Converting the list into a matrix
      ans <- do.call(rbind, ans)
      return(ans)
    }
    ```

2.  Once you have a version of the `my_boot()` function that runs on multiple cores, check that it provides accurate results by comparing it to a parametric model:

    ```{r}
    library (parallel)
    # Bootstrap of an OLS
    my_stat <- function(d) coef(lm(y ~ x, data = d))

    # Data Sim

    set.seed(1)
    n <- 500; R <- 1e4
    x <- cbind(rnorm(n)); y <- x * 5 + rnorm(n)

    # Checking if we get something similar as lm

    ans0 <- confint(lm(y ~ x))
    ans1 <- my_boot(dat = data.frame(x, y), my_stat, R = R, ncpus = 2L)

    t(apply(ans1, 2, quantile, c(.025, .975)))
    ans0

    ```

3.  Check whether your version actually goes faster when it’s run on multiple cores (since this might take a little while to run, we’ll use `system.time` and just run each version once, rather than `microbenchmark`, which would run each version 100 times, by default):

    ```{r}
    # Check speed
    system.time(my_boot(dat = data.frame(x, y), my_stat, R = R, ncpus = 1L))  
    system.time(my_boot(dat = data.frame(x, y), my_stat, R = R, ncpus = 2L))
    ```

### Problem 3: Short answer

1.  **Explain in your own words how you can decide if a problem is fit for parallelization?**

    Parallelization can be used when its tasks are independent of each other especially when one task does not need the result of the other to run and also if you have a large enough or computationally heavy task.

2.  **We have covered vectorization and parallelization in this lab. Explain in your own words how you can think about the relationship between vectorization and parallelization? How do we know whether to use one, the other, or both?**

    Vectorization means replacing slow R loops with fast built-in functions, while parallelization means splitting work across multiple CPU cores at the same time. I think vectorization is like just doing the same task faster on one core, but parallelization is splitting the jobs across multiple cores.

    Vectorization can be used if you are doing the same operation on many values which will make each step faster, and parallelization can be used if your task can be split into separate chunks that different processors can work on independently. You can also use both for instance vectorize first to make each tasks efficient and the parallelize to run multiple chunks of the task at the same time.
