---
title: "Lab 10. Interactive Visualization"
author: "Faith Okaalo"
format: 
  html:
    embed-resources: true
    code-fold: false
    toc: true
    toc-depth: 3
    theme: cosmo
editor: visual
---

## Lab 10: Interactive Visualization

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE)
```

Load the necessary packages

```{r}
library(dplyr)
library(ggplot2)
library(plotly)
library(zoo)
library(tidyr)
```

### Step 1: Read in the data

```{r}
# Load COVID state-level data from NYT
cv_states <- read.csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv")

# Load state population data
state_pops <- read.csv("https://raw.githubusercontent.com/COVID19Tracking/associated-data/master/us_census_data/us_census_2018_population_estimates_states.csv")
```

Check column names

```{r}
colnames(state_pops)
colnames(cv_states)
```

Adjust column names

```{r}
state_pops$abbs <- state_pops$state
state_pops$state <- state_pops$state_name
state_pops$state_name <- NULL

```

Merge the two datasets

```{r}
cv_states <- merge(cv_states, state_pops, by = "state")
```

### Step 2: Look at the data

```{r}
dim(cv_states)
```

```{r}
head(cv_states)
```

```{r}
tail(cv_states)
```

```{r}
str(cv_states)
```

### Step 3: Format the data 

```{r}
# Format the date
cv_states$date <-cv_states$date <- as.Date(cv_states$date, format = "%Y-%m-%d")

# Format state and state abbreviation (abb) variables
cv_states$state <- factor(cv_states$state)
cv_states$abb   <- factor(cv_states$abb)

# Order first by state, then by date
cv_states <- cv_states[order(cv_states$state, cv_states$date), ]

# Confirm formatting
str(cv_states)
head(cv_states)
tail(cv_states)

```

```{r}
# Inspect range values for each variable
head(cv_states)
summary(cv_states)


```

```{r}
min(cv_states$date)
```

```{r}
max(cv_states$date)
```

### Step 4: Add new_cases, new_deaths and correct outliers 

```{r}
state_list <- levels(cv_states$state)

# Initialize new columns
cv_states$new_cases  <- 0
cv_states$new_deaths <- 0

# Add variables for new_cases and new_deaths
for (i in 1:length(state_list)) {
  cv_subset <- subset(cv_states, state == state_list[i])
  cv_subset <- cv_subset[order(cv_subset$date), ]

  # Add starting level for new cases and deaths
  cv_subset$new_cases[1]  <- cv_subset$cases[1]
  cv_subset$new_deaths[1] <- cv_subset$deaths[1]

  for (j in 2:nrow(cv_subset)) {
    cv_subset$new_cases[j]  <- cv_subset$cases[j]  - cv_subset$cases[j - 1]
    cv_subset$new_deaths[j] <- cv_subset$deaths[j] - cv_subset$deaths[j - 1]
  }

  cv_states$new_cases[cv_states$state  == state_list[i]] <- cv_subset$new_cases
  cv_states$new_deaths[cv_states$state == state_list[i]] <- cv_subset$new_deaths
}

# Focus on recent dates
cv_states <- cv_states |> dplyr::filter(date >= "2021-06-01")
```

**Inspect outliers in new cases using plotly**

```{r}
p1 <- ggplot(cv_states, aes(x = date, y = new_cases, color = state)) +
  geom_line() +
  geom_point(size = .5, alpha = 0.5)
ggplotly(p1)

p2 <- ggplot(cv_states, aes(x = date, y = new_deaths, color = state)) +
  geom_line() +
  geom_point(size = .5, alpha = 0.5)
ggplotly(p2)
```

**Correct outliers**

```{r}
# set negative new cases or death counts to 0
cv_states$new_cases[cv_states$new_cases   < 0] <- 0
cv_states$new_deaths[cv_states$new_deaths < 0] <- 0

# Re-calculate cases and deaths as cumulative sum of updated new_cases / new_deaths
for (i in 1:length(state_list)) {
  cv_subset <- subset(cv_states, state == state_list[i])

  cv_subset$cases[1]  <- cv_subset$cases[1]
  cv_subset$deaths[1] <- cv_subset$deaths[1]

  for (j in 2:nrow(cv_subset)) {
    cv_subset$cases[j]  <- cv_subset$new_cases[j]  + cv_subset$cases[j - 1]
    cv_subset$deaths[j] <- cv_subset$new_deaths[j] + cv_subset$deaths[j - 1]
  }

  cv_states$cases[cv_states$state  == state_list[i]] <- cv_subset$cases
  cv_states$deaths[cv_states$state == state_list[i]] <- cv_subset$deaths
}
```

**Smooth new counts and inspect the data again**

```{r}
# Smooth new counts 
cv_states$new_cases  <- zoo::rollmean(cv_states$new_cases,  k = 7, fill = NA, align = 'right') |> round(digits = 0)
cv_states$new_deaths <- zoo::rollmean(cv_states$new_deaths, k = 7, fill = NA, align = 'right') |> round(digits = 0)

# Inspect data again interactively
p2 <- ggplot(cv_states, aes(x = date, y = new_deaths, color = state)) +
  geom_line() +
  geom_point(size = .5, alpha = 0.5)
ggplotly(p2)
```

### Step 5: Add additional variables 

```{r}
# Add population-normalized (per 100,000) counts
cv_states$per100k         <- as.numeric(format(round(cv_states$cases      / (cv_states$population / 100000), 1), nsmall = 1))
cv_states$newper100k      <- as.numeric(format(round(cv_states$new_cases  / (cv_states$population / 100000), 1), nsmall = 1))
cv_states$deathsper100k   <- as.numeric(format(round(cv_states$deaths     / (cv_states$population / 100000), 1), nsmall = 1))
cv_states$newdeathsper100k <- as.numeric(format(round(cv_states$new_deaths / (cv_states$population / 100000), 1), nsmall = 1))

# Add a naive CFR variable = deaths / cases
cv_states <- cv_states |> mutate(naive_CFR = round((deaths/cases),2))

# Create cv_states_today
cv_states_today <- subset(cv_states, date == max(cv_states$date))
```

### Step 6: Explore scatterplots using plot_ly

```{r}
# pop_density vs. cases
cv_states_today |>
  plot_ly(x = ~pop_density, y = ~cases,
          type = 'scatter', mode = 'markers', color = ~state,
          size = ~population, sizes = c(5, 70),
          marker = list(sizemode = 'diameter', opacity = 0.5))

# Filter out District of Columbia (outlier)
cv_states_today_filter <- cv_states_today |> filter(state != "District of Columbia")

# pop_density vs. cases after filtering
cv_states_today_filter |>
  plot_ly(x = ~pop_density, y = ~cases,
          type = 'scatter', mode = 'markers', color = ~state,
          size = ~population, sizes = c(5, 70),
          marker = list(sizemode = 'diameter', opacity = 0.5))

# pop_density vs. deathsper100k
cv_states_today_filter |>
  plot_ly(x = ~pop_density, y = ~deathsper100k,
          type = 'scatter', mode = 'markers', color = ~state,
          size = ~population, sizes = c(5, 70),
          marker = list(sizemode = 'diameter', opacity = 0.5))

# Adding hoverinfo
cv_states_today_filter |>
  plot_ly(x = ~pop_density, y = ~deathsper100k,
          type = 'scatter', mode = 'markers', color = ~state,
          size = ~population, sizes = c(5, 70),
          marker = list(sizemode = 'diameter', opacity = 0.5),
          hoverinfo = 'text',
          text = ~paste(
            paste(state, ":", sep = ""),
            paste(" Cases per 100k: ",  per100k,        sep = ""),
            paste(" Deaths per 100k: ", deathsper100k,  sep = ""),
            sep = "<br>"
          )) |>
  layout(
    title  = "Population-normalized COVID-19 deaths (per 100k) vs. population density for US states",
    yaxis  = list(title = "Deaths per 100k"),
    xaxis  = list(title = "Population Density"),
    hovermode = "compare"
  )
```

### Step 7: Explore scatterplot trend interactively using ggplotly() and geom_smooth()

```{r}
cv_states_today_scatter <- cv_states_today_filter 

p <- ggplot(cv_states_today_scatter,
            aes(x = pop_density, y = deathsper100k, size = population)) +
  geom_point() +
  geom_smooth()

ggplotly(p)
```

The scatterplot shows a weak, non-linear relationship between population density and COVID-19 deaths per 100,000. While the smoothed trend suggests a slight increase in death rates with higher population density, there is some substantial variability especially at the lower density levels.

### Step 8: Multiple line chart

```{r}
# Line chart for naive_CFR for all states over time
plot_ly(cv_states, x=~date, color=~state, y=~naive_CFR, type='scatter', mode='lines')
```

```{r}
# Line chart for Florida: new_cases and new_deaths together
cv_states |> filter(state == "Florida") |>
  plot_ly(x = ~date, y = ~new_cases,
          type = "scatter", mode = "lines",
          name = "New Cases") |>
  add_trace(x = ~date, y = ~new_deaths,
            type = "scatter", mode = "lines",
            name = "New Deaths")
```

The cases peaked in January 2022. The deaths are few and peaked in September 2021.

### Step 9: Heat maps

```{r}
# Map state, date, and new_cases to a matrix
cv_states_mat <- cv_states |>
  select(state, date, new_cases) |>
  dplyr::filter(date > as.Date("2021-06-15"))

cv_states_mat2 <- as.data.frame(
  pivot_wider(cv_states_mat, names_from = state, values_from = new_cases)
)
rownames(cv_states_mat2) <- cv_states_mat2$date
cv_states_mat2$date      <- NULL
cv_states_mat2           <- as.matrix(cv_states_mat2)

# Heatmap of new_cases
plot_ly(x = colnames(cv_states_mat2), y = rownames(cv_states_mat2),
        z = ~cv_states_mat2,
        type = "heatmap",
        showscale = TRUE)
```

States such as California, Texas, Florida, and New York had higher cases of COVID, especially at the beginning of 2022.

**Repeat with newper100k**

```{r}
cv_states_mat <- cv_states |>
  select(state, date, newper100k) |>
  dplyr::filter(date > as.Date("2021-06-15"))

cv_states_mat2 <- as.data.frame(
  pivot_wider(cv_states_mat, names_from = state, values_from = newper100k)
)
rownames(cv_states_mat2) <- cv_states_mat2$date
cv_states_mat2$date      <- NULL
cv_states_mat2           <- as.matrix(cv_states_mat2)

plot_ly(x = colnames(cv_states_mat2), y = rownames(cv_states_mat2),
        z = ~cv_states_mat2,
        type = "heatmap",
        showscale = TRUE)

```

The previous map showed that the big states, such as California and Texas had high cases of COVID may be due to their population size. The heatmap with cases per 100,000 population shows that there was a spike in cases across all states.

```{r}
# Heatmap filtered to every two weeks
filter_dates <- seq(as.Date("2021-06-15"), as.Date("2021-11-01"), by = "2 weeks")

cv_states_mat <- cv_states |>
  select(state, date, newper100k) |>
  filter(date %in% filter_dates)

cv_states_mat2 <- as.data.frame(
  pivot_wider(cv_states_mat, names_from = state, values_from = newper100k)
)
rownames(cv_states_mat2) <- cv_states_mat2$date
cv_states_mat2$date      <- NULL
cv_states_mat2           <- as.matrix(cv_states_mat2)

plot_ly(x = colnames(cv_states_mat2), y = rownames(cv_states_mat2),
        z = ~cv_states_mat2,
        type = "heatmap",
        showscale = TRUE)
```

The heatmap of COVID-19 cases per 100,000 population, filtered to biweekly intervals, reveals a clear temporal trend from June to November 2021. Case rates were relatively low during early summer but increased across most states from August. There were differences across states, with some having higher case rates than others.

### Step 10. Map

```{r}
pick.date <- "2021-10-15"

# Extract data for each state by its abbreviation
cv_per100 <- cv_states |>
  filter(date == pick.date) |>
  select(state, abb, newper100k, cases, deaths)

cv_per100$state_name <- cv_per100$state
cv_per100$state      <- cv_per100$abb
cv_per100$abb        <- NULL

# Create hover text
cv_per100$hover <- with(cv_per100,
  paste(state_name, '<br>',
        "Cases per 100k: ", newper100k, '<br>',
        "Cases: ",          cases,      '<br>',
        "Deaths: ",         deaths))

# Set up mapping details
set_map_details <- list(
  scope      = 'usa',
  projection = list(type = 'albers usa'),
  showlakes  = TRUE,
  lakecolor  = toRGB('white')
)

# Shared color scale limit
shadeLimit <- 125

# Map for Oct 15, 2021
fig <- plot_geo(cv_per100, locationmode = 'USA-states') |>
  add_trace(
    z         = ~newper100k,
    text      = ~hover,
    locations = ~state,
    color     = ~newper100k,
    colors    = 'Purples'
  )
fig <- fig |> colorbar(title  = paste0("Cases per 100k: ", pick.date),
                       limits = c(0, shadeLimit))
fig <- fig |> layout(
  title = paste('Cases per 100k by State as of', pick.date, '<br>(Hover for value)'),
  geo   = set_map_details
)
fig_pick.date <- fig
fig_pick.date
```

```{r}
# Map for most recent date
cv_per100 <- cv_states_today |>
  select(state, abb, newper100k, cases, deaths)

cv_per100$state_name <- cv_per100$state
cv_per100$state      <- cv_per100$abb
cv_per100$abb        <- NULL

# Create hover text
cv_per100$hover <- with(cv_per100,
  paste(state_name, '<br>',
        "Cases per 100k: ", newper100k, '<br>',
        "Cases: ",          cases,      '<br>',
        "Deaths: ",         deaths))

# Set up mapping details
set_map_details <- list(
  scope      = 'usa',
  projection = list(type = 'albers usa'),
  showlakes  = TRUE,
  lakecolor  = toRGB('white')
)

fig <- plot_geo(cv_per100, locationmode = 'USA-states') |>
  add_trace(
    z         = ~newper100k,
    text      = ~hover,
    locations = ~state,
    color     = ~newper100k,
    colors    = 'Purples'
  )
fig <- fig |> colorbar(title  = paste0("Cases per 100k: ", Sys.Date()),
                       limits = c(0, shadeLimit))
fig <- fig |> layout(
  title = paste('Cases per 100k by State as of', Sys.Date(), '<br>(Hover for value)'),
  geo   = set_map_details
)
fig_Today <- fig
fig_Today
```

**Plot the maps together**

```{r}
subplot(fig_pick.date, fig_Today, nrows = 2, margin = .05)
```

Compared with October 15, 2021, the most recent map shows very few reported COVID-19 cases, with uniformity across the country. On October 15, 2021, several states had high case counts and stand out in darker shading.
