---
title: "Lab 9. SQL"
author: "Faith Okaalo"
date: "03/17/2026"
format: html
editor: visual
---

### Load the required packages

```{r}
library(RSQLite)
library(DBI)
```

### Initialize a temporary in memory database

```{r}
con <- dbConnect(SQLite(), ":memory:")

```

### Download tables

```{r}
doctor <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/doctor.csv")
visit <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/visit.csv")
patient <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/patient.csv")
payment <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/payment.csv")


# Copy data frames to database

dbWriteTable(con, "doctor", doctor)
dbWriteTable(con, "visit", visit)
dbWriteTable(con, "patient", patient)
dbWriteTable(con, "payment", payment)

dbListTables(con)
```

### Check the table structure

```{r pragma}
dbGetQuery(con, "PRAGMA table_info(doctor)")
```

### Exercise 1: Retrieve the doctor ID, first name, and last name for all doctors. Sort by last name then first name.

```{sql connection=con}
SELECT doctor_id, first_name, last_name
FROM doctor
ORDER BY last_name, first_name
```

### Exercise 2: Retrieve the doctor ID, first name, and last name for doctors whose last name is 'WILLIAMS' or 'DAVIS'

```{sql connection=con}
SELECT doctor_id, first_name, last_name
FROM doctor
WHERE last_name IN ('WILLIAMS', 'DAVIS')
```

### Exercise 3: Distinct patient IDs who visited the hospital on July 5, 2005.

```{sql connection=con}
SELECT DISTINCT patient_id
FROM visit
WHERE date(visit_start_date) = '2005-07-05'
```

### Exercise 4

##### 4.1 Payment rows where the amount is 199, 799, or 999.

```{sql connection=con}
SELECT *
FROM payment
WHERE amount IN (199, 799, 999)
```

##### 4.2 Payment rows where the amount is greater than 500

```{sql connection=con}
SELECT *
FROM payment
WHERE amount > 500
```

##### 4.3 Payment rows where the amount is greater than 500 and less than 800.

```{sql connection=con}
SELECT *
FROM payment
WHERE amount > 500 AND amount < 800

```

### Exercise 5: Retrieve all payment IDs and their amounts from patients whose last name is 'DAVIS'

```{sql connection=con}
SELECT payment.payment_id, payment.amount
FROM payment
  INNER JOIN patient ON payment.patient_id = patient.patient_id
WHERE patient.last_name = 'DAVIS'
```

#### Exercise 6:

##### 6.1 Use `COUNT(*)` to count the number of rows in `visit`.

```{sql connection=con}
SELECT COUNT(*) AS total_visits
FROM visit
```

##### 6.2 Use `COUNT(*)` and `GROUP BY` to count the number of visits for each `patient_id`.

```{sql connection=con}
SELECT patient_id, COUNT(*) AS num_visits
FROM visit
GROUP BY patient_id
```

##### 6.3 Repeat the previous query and sort by the count in descending order.

```{sql connection=con}
SELECT patient_id, COUNT(*) AS num_visits
FROM visit
GROUP BY patient_id
ORDER BY num_visits DESC
```

##### 6.4 Repeat the previous query but use `HAVING` to only keep the groups with 40 or more.

```{sql connection=con}
SELECT patient_id, COUNT(*) AS num_visits
FROM visit
GROUP BY patient_id
HAVING num_visits >= 40
ORDER BY num_visits DESC
```

####  **Exercise 7**

**Write a query that calculates a number of summary statistics for `amount` in the payment table using `MAX`, `MIN`, `AVG` and `SUM`**

```{sql connection=con}
SELECT
  MAX(amount) AS max_amount,  
  MIN(amount) AS min_amount,  
  AVG(amount) AS avg_amount,  
  SUM(amount) AS sum_amount   
FROM payment
```

##### 7.1 Modify the above query to do those calculations for each `patient_id`.

```{sql connection=con}
SELECT
  patient_id,
  MAX(amount) AS max_amount,
  MIN(amount) AS min_amount,
  AVG(amount) AS avg_amount,
  SUM(amount) AS sum_amount,
  COUNT(*) AS num_payments
FROM payment
GROUP BY patient_id
```

##### 7.2 Modify the above query to only keep the `patient_id`s that have more then 5 payments

```{sql connection=con}
SELECT
  patient_id,
  MAX(amount) AS max_amount,
  MIN(amount) AS min_amount,
  AVG(amount) AS avg_amount,
  SUM(amount) AS sum_amount,
  COUNT(*) AS num_payments 
FROM payment
GROUP BY patient_id
HAVING num_payments > 5

```

#### Exercise 8.1 Create a SQL query to join the `doctor` and `visit` tables. Then, use SQL to compute the average visit length for each doctor. Hint: use the SQLite function `julianday` to convert a text date into a number.

```{sql connection=con}
SELECT
  doctor.doctor_id,
  doctor.first_name,
  doctor.last_name,
  AVG(julianday(visit_end_date) - julianday(visit_start_date)) AS avg_visit_length
FROM visit
  INNER JOIN doctor ON visit.doctor_id = doctor.doctor_id
GROUP BY doctor.doctor_id
```

#### Exercise 8.2 What are the names of the doctors with the 5 longest and 5 shortest average visits? Hint: use the `LIMIT` command.

**Doctors with the 5 longest average visit lengths.**

```{sql connection=con}
SELECT
  doctor.first_name,
  doctor.last_name,
  AVG(julianday(visit_end_date) - julianday(visit_start_date)) AS avg_visit_length
FROM visit
  INNER JOIN doctor ON visit.doctor_id = doctor.doctor_id
GROUP BY doctor.doctor_id
ORDER BY avg_visit_length DESC
LIMIT 5
```

**Doctors with the 5 shortest average visit lengths.**

```{sql connection=con}
SELECT
  doctor.first_name,
  doctor.last_name,
  AVG(julianday(visit_end_date) - julianday(visit_start_date)) AS avg_visit_length
FROM visit
  INNER JOIN doctor ON visit.doctor_id = doctor.doctor_id
GROUP BY doctor.doctor_id
ORDER BY avg_visit_length ASC
LIMIT 5
```

**Cleanup**

```{r}
dbDisconnect(con)
```
