---
title: "LAB 5. DATA WRANGLING"
author: "Faith Okaalo"
date: 02/12/2026
format: 
  html:
    embed-resources: true
    code-fold: false
    toc: true
    toc-depth: 3
    theme: cosmo
editor: visual
---

### Step 1: Load libraries

```{r}
# Load required packages
library(data.table)
library(dplyr)
library(tidyr)
library(janitor)
library(jsonlite)
library(httr2)
library(lubridate)
```

### Step 2: Read in the data

```{r}
url <- 'https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/covid/population.csv'
if (!file.exists("population.csv"))
  download.file(
    url = url,
    destfile = "population.csv",
    method   = "libcurl",
    timeout  = 60
  )
population <- read.csv('population.csv')
```

Print the first few rows

```{r}
print(head(population))
```

### Step 3: Examine the population matrix

```{r}
# Use janitor to make first row the header
population <- population %>%
  row_to_names(row_number = 1)
```

```{r}
names(population)
print(head(population))
```

```{r}
# Remove unnecessary columns (index and state ID)
population <- population %>%
  select(-1, -state)
```

```{r}
# Rename state column
population <- population %>%
  rename(state_name = NAME)
```

```{r}
# Convert POP_2020 and POP_2021 columns to numeric because they are saved as characters
population <- population %>%
  mutate(
    POP_2020 = as.numeric(POP_2020),
    POP_2021 = as.numeric(POP_2021)
  )
```

```{r}
# Add state abbreviations
state_lookup <- c(
  setNames(state.abb, state.name),
  "District of Columbia" = "DC",
  "Puerto Rico" = "PR")
```

```{r}
population <- population %>%
  mutate(state = state_lookup[state_name])

# Print first few rows
head(population)
```

```{r}
# Convert to tidy format
population <- population %>%
  pivot_longer(
    cols = starts_with("POP_"),
    names_to = "year",
    values_to = "population"
  ) %>%
  mutate(year = as.integer(gsub("POP_", "", year)))

# Print first few rows
head(population)
```

### Step 4: Load regions data

```{r}
url <- "https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/covid/regions.json"
regions <- fromJSON(url)
```

```{r}
# Split states into separate rows using unnest
regions <- regions %>%
  unnest(states) %>%
  rename(state_name = states)
```

```{r}
# Make region a factor
regions <- regions %>%
  mutate(region = as.character(region)) %>%
  mutate(region = as.factor(region))
```

```{r}
# Shorten long region name
regions <- regions %>%
  mutate(region_name = ifelse(region_name == "New York and New Jersey, Puerto Rico, Virgin Islands",
                               "NY-NJ-PR-VI",
                               region_name))

print(regions)
```

### Step 5: Add a region and region name columns to the population data frame.

```{r}
population <- population %>%
  left_join(regions, by = "state_name")

# Print first few rows
head(population)
```

### Step 6: CDC COVID cases data

```{r}
api <- "https://data.cdc.gov/resource/pwn4-m3yp.json"
request <- request(paste0(api, paste0("?$limit=10000000000")))
response <- request |> req_perform() |> resp_body_string()
cases_raw <- fromJSON(response)

# Print first few rows
head(cases_raw)
```

\# Answer if we replace "?\$limit=10000000000" with "", the API returns a limited number of records not the entire dataset so we end up having an incomplete dataset.

### Step 7: Wrangle cases data

```{r}
cases <- cases_raw %>%
  select(state = state, 
         date = end_date, 
         cases = new_cases) %>%
  mutate(cases = as.numeric(cases)) %>%
  mutate(date = ymd_hms(date)) %>%
  mutate(date = as_date(date))

# Print first few rows
head(cases)
```

### Step 8: Merge with population data

```{r}
cases_full <- cases %>%
  mutate(year = year(date)) %>%
  left_join(population, by = c("state", "year"))

# Add epi week and year
cases_full <- cases_full %>%
  mutate(
    epi_week = epiweek(date),
    epi_year = epiyear(date)
  )

# Order by date within state
cases_full <- cases_full %>%
  arrange(state, date)

# Print first few rows
head(cases_full)
```
